Industrial, large-scale model predictive control with structured neural networks

### 2021-02-05 paper.tex
add source code section.
update paper.tex

%% This is file `elsarticle-template-1-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1-num.tex 149 2009-10-08 05:01:15Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1-num.tex $
%%
\documentclass[preprint,5p, twocolumn, authoryear]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% The graphicx package provides the includegraphics command.
\usepackage{graphicx}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amsmath}   % Extra math commands and environments from the AMS
\usepackage{amssymb}   % Special symbols from the AMS
\usepackage{amsthm}    % Enhanced theorem and proof environments from the AMS
\usepackage{latexsym}  % A few extra LaTeX symbols
\usepackage{array}
\usepackage{makecell}
\usepackage{units}
\usepackage{xcolor}
%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
\usepackage{lineno}
\usepackage[singlelinecheck=false]{caption}

@@

%\biboptions{longnamesfirst, semicolon}
%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

@@

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
\author[label1]{Pratyush Kumar\corref{cor1}}
\ead{pratyushkumar@ucsb.edu}
%% \ead[url]{home page}
%% \fntext[label2]{}
\cortext[cor1]{Corresponding Author}
%% \fntext[label3]{}

@@

\subsection{Structured neural network}
An intuitive approach is to build a feedforward NN that takes the triple ($x,
x_s, u_s$) as its input and outputs a control that is close to the optimal
control. This strategy has been proposed in \cite*{karg:lucia:2020} and 
\cite*{chen:saulnier:atanasov:lee:kumar:pappas:morari:2018}
for a single fixed steady state. The approach has not been demonstrated to scale
to the size of model dimensions and the setpoint tracking offset-free MPC
problem considered in this article. An issue with this strategy is that the NN
has no knowledge about the MPC feedback law, and can require large amounts of
training data in large state dimensions to obtain a reasonable controller. At a
minimum, the MPC feedback law is such that at the origin, the control action
taken by the MPC controller is zero. For the setpoint tracking MPC problem
considered in this article, the NN control law must be such that at steady-state
operations, the NN maintains the plant at the desired steady state such that the
controlled measurements are at their setpoints. This structure implies $u_s =
\kappa_{NN}(x = x_s, x_s, u_s)$, in which we use $\kappa_{NN}(\cdot)$ to denote
the NN control law. To incorporate this information in a NN, we introduce the
following architecture
\begin{align}
z_0 &= \begin{bmatrix}
    x', & x_s', & u_s', & x_s', & x_s', & u_s' \end{bmatrix}'\notag \\
    f_i(z_{i-1}) &= \begin{bmatrix}
        W_i & 0 \notag \\
        0 & W_i
          \end{bmatrix}z_{i-1}  + \begin{bmatrix}
            b_i \\
            b_i
            \end{bmatrix} \notag\\
    z_i &= g(f_i(z_{i-1})), \textnormal{for} \ i \in \bbI_{1:h} \notag \\
    u &= u_s + \begin{bmatrix}
  W_{h + 1}, & -W_{h + 1} \\
    \end{bmatrix} z_h
\label{eq:NNstruct}
\end{align}
in which $g(a) = \textnormal{max} (0, a)$, is the ReLU operation applied
element-wise on its vector input, $z_i$ is the output of each hidden layer of
the network, $z_0$ is the input to the network, $h$ is the number of hidden
layers, and $i \in \bbI_{1:h}$. The parameters to be optimized using training
data are ($W_i, b_i$) and $W_{h+1}$, which are matrices of appropriate
dimensions. At steady state, when $x=x_s$ in $z_0$, \eqref{eq:NNstruct} outputs
$u_s$ regardless of the choice of weights. The software tensorflow
\citep{abadi:agarwal:barham:brevdo:et-al:2015} is used for NN training purposes
in this article, and the structured architecture can be built with symbolic
differentiation tools available in tensorflow. When a rate-of-change penalty on
the control is included in the regulator QP, the input to the structured NN is
modified to 
\begin{align*}
    z_0 &= 
    \begin{bmatrix}
    x', & u_{-1}', & x_s', & u_s', & x_s', & u_s', & x_s', & u_s' \\
    \end{bmatrix}'
\end{align*}
in which $u_{-1}$ is the previous control applied to the plant and the
dimensions of the matrices ($W_i, b_i$) and $W_{h+1}$ are appropriately
adjusted. Even after training, the control produced by the NN may not satisfy
the hard actuator constraints, and the saturation function
($\textnormal{sat}(\cdot, \underline{u}, \overline{u})$) is applied to the
output of the NN.

@@

Figure \ref{fig:cdu_statistics} shows the statistics for the simulation study
performed for this example. The data requirement to obtain an accurate NN
controller is shown in the top left. The batch size in Adam was chosen as 2048
samples, and all the NNs are trained for 1500 epochs. For all the NN training,
the size of the buffer dataset was chosen as 5$\%$ of the training dataset. The
transient controller performance index for one validation simulation obtained
with the three structured NNs, satK, and MPC controllers is shown in the center
bottom. The top right plot shows the histogram of online computation times of
the NNs and QP solver. Table \ref{table:cdu} gives a summary of the maximum time
required to train the NNs, number of parameters in each NN, memory footprint of
the NNs, and the $\%$ performance loss metric obtained by the structured NNs,
SS, satK, and SH controllers in the validation simulations.

@@

\section{Conclusion} \label{sec:conclusion}
Neural networks have been demonstrated to approximate the MPC feedback law for
treating large-scale, linear MPC applications that may be out of reach with QP
solvers. The proposed methods avoiding online optimization to date in the
literature, such as storing the entire MPC feedback law and sampling the large
dimensional state space for the NN training, are not tractable for the problems
considered in this article.

The key features for the scalability of the proposed approach are the structure
of the NN and data generation for only typical plant operational scenarios. The
NN design approach was demonstrated to scale on an industrial crude distillation
unit model with state and control dimensions of 252 and 32 respectively, and a
control-sample horizon length of 140. After the offline design phase, NNs
execute MPC about three to five orders of magnitude faster than an available QP
solver with less than 1$\%$ performance degradation in the closed-loop. The
deployment decision of the NNs must be based on the ability to reliably identify
and sample the anticipated plant operational scenarios with setpoints and large
magnitude disturbances that may change for a reasonable duration of operation.
Parallel computing can be used for the offline data generation step if solving
QPs for each sampled state starts to take excessive amounts of time. After the
training process, the NNs are easily deployable on memory constrained hardware.
We established conditions under which the NN controllers are robust to state
estimation errors and process disturbances, provides some theoretical support to
practitioners interested in the deployment of NNs.

Future research in the area of approximating MPC control laws using NNs can
focus on developing methods for approximating the target selector QP solution,
systematically handling state constraints, extensions to nonlinear MPC problems,
and establishing conditions for guaranteed closed-loop stability with NN
controllers.

@@

>>>>

%% This is file `elsarticle-template-1-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public %
%License, either version 1.2 of this license or (at your option) any % later
%version.  The latest version of this license is in %
%http://www.latex-project.org/lppl.txt % and version 1.2 or later is part of all
%distributions of LaTeX % version 1999/12/01 or later.
%%
%% Template article for Elsevier's document class `elsarticle' % with numbered
%style bibliographic references
%%
%% $Id: elsarticle-template-1-num.tex 149 2009-10-08 05:01:15Z rishi $ % $URL:
%http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1-num.tex $
%%
\documentclass[preprint,5p, twocolumn, authoryear]{elsarticle}

%% Use the option review to obtain double line spacing %
%\documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn % for a
%journal layout: % \documentclass[final,1p,times]{elsarticle} %
%\documentclass[final,1p,times,twocolumn]{elsarticle} %
%\documentclass[final,3p,times]{elsarticle} %
%\documentclass[final,3p,times,twocolumn]{elsarticle} %
%\documentclass[final,5p,times]{elsarticle} %
%\documentclass[final,5p,times,twocolumn]{elsarticle}

%% The graphicx package provides the includegraphics command.
\usepackage{graphicx}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amsmath}   % Extra math commands and environments from the AMS
\usepackage{amssymb}   % Special symbols from the AMS
\usepackage{amsthm}    % Enhanced theorem and proof environments from the AMS
\usepackage{latexsym}  % A few extra LaTeX symbols
\usepackage{array}
\usepackage{makecell}
\usepackage{units}
\usepackage{xcolor}
\usepackage[hidelinks=true]{hyperref}

%% The lineno packages adds line numbers. Start line numbering with %
%\begin{linenumbers}, end it with \end{linenumbers}. Or switch it on % for the
%whole article with \linenumbers after \end{frontmatter}.
\usepackage{lineno}
\usepackage[singlelinecheck=false]{caption}

@@

%\biboptions{longnamesfirst, semicolon} % natbib.sty is loaded by default.
%However, natbib options can be % provided with \biboptions{...} command.
%Following options are % valid:

%%   round  -  round parentheses are used (default) %   square -  square
%brackets are used   [option] %   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option> %   semicolon  -  multiple
%citations separated by semi-colon %   colon  - same as semicolon, an earlier
%confusion %   comma  -  separated by comma %   numbers-  selects numerical
%citations %   super  -  numerical citations as superscripts %   sort   -  sorts
%multiple citations according to order in ref. list %   sort&compress   -  like
%sort, but also compresses numerical citations %   compress - compresses without
%sorting
%%
%% \biboptions{comma,round}

@@

%% use the tnoteref command within \title for footnotes; % use the tnotetext
%command for the associated footnote; % use the fnref command within \author or
%\address for footnotes; % use the fntext command for the associated footnote; %
%use the corref command within \author for corresponding author footnotes; % use
%the cortext command for the associated footnote; % use the ead command for the
%email address, % and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}} % \tnotetext[label1]{}
\author[label1]{Pratyush Kumar\corref{cor1}}
\ead{pratyushkumar@ucsb.edu}
%% \ead[url]{home page} % \fntext[label2]{}
\cortext[cor1]{Corresponding Author}
%% \fntext[label3]{}

@@

\subsection{Structured neural network}
An intuitive approach is to build a feedforward NN that takes the triple ($x,
x_s, u_s$) as its input and outputs a control that is close to the optimal
control. This strategy has been proposed in \cite*{karg:lucia:2020} and 
\cite*{chen:saulnier:atanasov:lee:kumar:pappas:morari:2018}
for a single fixed steady state. The approach has not been demonstrated to scale
to the size of model dimensions and the setpoint tracking offset-free MPC
problem considered in this article. An issue with this strategy is that the NN
has no knowledge about the MPC feedback law, and can require large amounts of
training data in large state dimensions to obtain a reasonable controller. At a
minimum, the MPC feedback law is such that at the origin, the control action
taken by the MPC controller is zero. For the setpoint tracking MPC problem
considered in this article, the NN control law must be such that at steady-state
operations, the NN maintains the plant at the desired steady state such that the
controlled measurements are at their setpoints. This structure implies $u_s =
\kappa_{NN}(x = x_s, x_s, u_s)$, in which we use $\kappa_{NN}(\cdot)$ to denote
the NN control law. To incorporate this information in a NN, we introduce the
following architecture
\begin{align}
z_0 &= \begin{bmatrix}
    x', & x_s', & u_s', & x_s', & x_s', & u_s' \end{bmatrix}'\notag \\
    f_i(z_{i-1}) &= \begin{bmatrix}
        W_i & 0 \notag \\
        0 & W_i
          \end{bmatrix}z_{i-1}  + \begin{bmatrix}
            b_i \\
            b_i
            \end{bmatrix} \notag\\
    z_i &= g(f_i(z_{i-1})), \textnormal{for} \ i \in \bbI_{1:h} \notag \\
    u &= u_s + \begin{bmatrix}
  W_{h + 1}, & -W_{h + 1} \\
    \end{bmatrix} z_h
\label{eq:NNstruct}
\end{align}
in which $g(a) = \textnormal{max} (0, a)$, is the ReLU operation applied
element-wise on its vector input, $z_i$ is the output of each hidden layer of
the network, $z_0$ is the input to the network, $h$ is the number of hidden
layers, and $i \in \bbI_{1:h}$. The parameters to be optimized using training
data are ($W_i, b_i$) and $W_{h+1}$, which are matrices of appropriate
dimensions. At steady state, when $x=x_s$ in $z_0$, \eqref{eq:NNstruct} outputs
$u_s$ regardless of the choice of weights as the output of all hidden layers
($z_i$) have same subvector repeated in the upper and lower half and
$\begin{bmatrix} W_{h + 1}, & -W_{h+ 1} \\
      \end{bmatrix} z_h$ equals zero. The software tensorflow
\citep{abadi:agarwal:barham:brevdo:et-al:2015} is used for NN training purposes
in this article, and the structured architecture can be built with symbolic
differentiation tools available in tensorflow. When a rate-of-change penalty on
the control is included in the regulator QP, the input to the structured NN is
modified to 
\begin{align*}
    z_0 &= 
    \begin{bmatrix}
    x', & u_{-1}', & x_s', & u_s', & x_s', & u_s', & x_s', & u_s' \\
    \end{bmatrix}'
\end{align*}
in which $u_{-1}$ is the previous control applied to the plant and the
dimensions of the matrices ($W_i, b_i$) and $W_{h+1}$ are appropriately
adjusted. Even after training, the control produced by the NN may not satisfy
the hard actuator constraints, and the saturation function
($\textnormal{sat}(\cdot, \underline{u}, \overline{u})$) is applied to the
output of the NN.

@@

Figure \ref{fig:cdu_statistics} shows the statistics for the simulation study
performed for this example. The data requirement to obtain an accurate NN
controller is shown in the top left. The batch size in Adam was chosen as 2048
samples, and all the NNs are trained for 1500 epochs. The size of the buffer
dataset was chosen as 5$\%$ of the training dataset. The transient controller
performance index for one validation simulation obtained with the three
structured NNs, satK, and MPC controllers is shown in the center bottom. The top
right plot shows the histogram of online computation times of the NNs and QP
solver. Table \ref{table:cdu} gives a summary of the maximum time required to
train the NNs, number of parameters in each NN, memory footprint of the NNs, and
the $\%$ performance loss metric obtained by the structured NNs, SS, satK, and
SH controllers in the validation simulations.

@@

\section{Conclusion} \label{sec:conclusion}

Neural networks have been demonstrated to approximate the MPC feedback law for
treating large-scale, linear MPC applications that may be out of reach with QP
solvers. The proposed methods avoiding online optimization to date in the
literature, such as storing the entire MPC feedback law and sampling the large
dimensional state space for the NN training, are not tractable for the problems
considered in this article.

The key features for the scalability of the proposed approach are the structure
of the NN and data generation for only typical plant operational scenarios. The
NN design approach was demonstrated to scale on an industrial crude distillation
unit model with state and control dimensions of 252 and 32 respectively, and a
control-sample horizon length of 140. After the offline design phase, NNs
execute MPC about three to five orders of magnitude faster than an available QP
solver with less than 1$\%$ performance degradation in the closed-loop. The
deployment decision of the NNs must be based on the ability to reliably identify
and sample the anticipated plant operational scenarios with setpoints and large
magnitude disturbances that may change for a reasonable duration of operation.
Parallel computing can be used for the offline data generation step if solving
QPs for each sampled state starts to take excessive amounts of time. After the
training process, the NNs are easily deployable on memory constrained hardware.
We established conditions under which the NN controllers are robust to state
estimation errors and process disturbances, provides some theoretical support to
practitioners interested in the deployment of NNs. Future research may be
focused to establish conditions for guaranteed closed-loop stability with NN
controllers.

@@

\section*{Source code}
The code for the simulation studies performed in this article can be found at:
\url{https://github.com/pratyushkumar211/industrial_nnmpc_2021}
